---
title: "Retail Data Analysis: Sales Analysis and Customer Segmentation"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
# Load necessary libraries
library(dplyr)         
library(ggplot2)        
library(lubridate)     
library(treemap)        
library(cluster)      
library(scales)
library(readxl)
library(tidyr)
options(warn = -1)  # Suppress warnings for cleaner output
```

1. Load Data

```{r}
# Read the Excel file
df <- read_excel("C:/Users/Wenwen/Desktop/case study/online_retail_II.xlsx")
head(df)  # Display the first few rows of the dataset
```

2. Data Discovery
Summary Statistics

```{r}
summary(df) 
```

Data Structure

```{r}
str(df) 
```

Unique Products

```{r}
n_distinct(df$Description)  
```

Missing Values

```{r}
colSums(is.na(df)) 
```

Unique Customers
```{r}
```

3. Data Cleaning

Remove Duplicates

```{r}
sum(duplicated(df))  
df_no_duplicates <- df %>% distinct()  
cat("Number of rows after removing duplicates:", nrow(df_no_duplicates), "\n")
cat("Number of rows before removing duplicates:", nrow(df), "\n")
```

Handle Missing Values

```{r}
colSums(is.na(df_no_duplicates))  
```

Fill Missing Descriptions

```{r}
# Function to get the mode of a vector
get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Fill missing descriptions based on the mode of the StockCode
df_no_duplicates <- df_no_duplicates %>%
  group_by(StockCode) %>%
  mutate(Description = ifelse(is.na(Description), get_mode(Description), Description)) %>%
  ungroup()

# Replace any remaining NA descriptions with "No Description"
df_no_duplicates <- df_no_duplicates %>%
  mutate(Description = replace_na(Description, "No Description"))

# Check for missing values again
colSums(is.na(df_no_duplicates))
```

Handle Missing Customer IDs

```{r}
# Replace missing Customer IDs with '-1' for type conversion
df_no_duplicates <- df_no_duplicates %>%
  mutate(`Customer ID` = ifelse(is.na(`Customer ID`), -1, `Customer ID`))

head(df_no_duplicates)
```
For null values in CustomerID, we replace them with '-1' to enable type conversion, as the current type cannot handle null values.

For the Description column, we fill missing values by mapping the StockCode to its corresponding description, and set 'No Description' for any remaining null values.


4. Data Manipulation
Extract Time-Based Features

```{r}
# Extract month, week, and day from InvoiceDate
df_no_duplicates <- df_no_duplicates %>%
  mutate(
    Month = month(InvoiceDate),
    MonthText = month(InvoiceDate, label = TRUE, abbr = TRUE),
    Week = week(InvoiceDate),
    Day = day(InvoiceDate)
  )

head(df_no_duplicates)
```

5. Exploratory Data Analysis (EDA)
Summary of Cleaned Data

```{r}
summary(df_no_duplicates)  
```

Boxplot of Quantity

```{r}
boxplot(df_no_duplicates$Quantity, main = "Boxplot of Quantity")
```

Boxplot of Price

```{r}
boxplot(df_no_duplicates$Price, main = "Boxplot of Price")
```

5.2 Sales Trend by Time

Calculate Sales

```{r}
# Calculate Sales as the product of Quantity and Price
df_no_duplicates <- df_no_duplicates %>%
  mutate(Sales = Quantity * Price)
```

Monthly Sales Trend

```{r}
# Group by Month and calculate total sales
monthly_sales <- df_no_duplicates %>%
  group_by(Month) %>%
  summarize(TotalSales = sum(Sales))

# Define month names for plotting
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Plot the total sales trend by month
ggplot(monthly_sales, aes(x = factor(Month, labels = months), y = TotalSales)) +
  geom_point(size = 3) +  # Use points to represent sales
  geom_line(group = 1) +  # Connect points with lines
  labs(title = "Total Sales Trend by Month",
       x = "Month",
       y = "Total Sales") +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "gray80"))
```

Weekly Sales Trend

```{r}
# Group by Week and calculate total sales
weekly_sales <- df_no_duplicates %>%
  group_by(Week) %>%
  summarize(TotalSales = sum(Sales))

# Plot the sales trend by week using an area plot
ggplot(weekly_sales, aes(x = Week, y = TotalSales)) +
  geom_area(fill = "steelblue", alpha = 0.5) +  # Create an area plot
  labs(title = "Sales Trend by Week",
       x = "Week",
       y = "Total Sales") +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "gray80"))
```

Daily Sales Trend
```{r}
# Extract day of the week from InvoiceDate
df_no_duplicates <- df_no_duplicates %>%
  mutate(DayOfWeek = wday(InvoiceDate, label = TRUE, abbr = TRUE))

# Group by DayOfWeek and calculate total sales
day_of_week_sales <- df_no_duplicates %>%
  group_by(DayOfWeek) %>%
  summarize(TotalSales = sum(Sales))

# Plot the total sales trend by day of the week
ggplot(day_of_week_sales, aes(x = DayOfWeek, y = TotalSales)) +
  geom_point(size = 3) +  # Use points to represent sales
  geom_line(group = 1) +  # Connect points with lines
  labs(title = "Total Sales Trend by Day of Week",
       x = "Day of Week",
       y = "Total Sales") +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "gray80"))
```
The highest sales occur between September and December, peaking in November.
The week with the largest sales is week 49, just before the Christmas holiday.
Customers make the fewest purchases on weekends, with almost no sales on Saturdays.

5.3 Sales by Product
Top 10 Products by Sales

```{r}
# Group by Description, calculate total sales, and get the top 10 products
product_sales <- df_no_duplicates %>%
  group_by(Description) %>%
  summarize(TotalSales = sum(Sales)) %>%
  arrange(desc(TotalSales)) %>%
  head(10)

# Plot the top 10 products by sales
ggplot(product_sales, aes(x = TotalSales, y = reorder(Description, TotalSales))) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Create a horizontal bar plot
  labs(title = "Top 10 Products by Sales",
       x = "Total Sales",
       y = "Product") +
  theme_minimal()
```

Top 10 Products by Quantity

```{r}
# Group by Description, calculate total quantities, and get the top 10 products
product_quantities <- df_no_duplicates %>%
  group_by(Description) %>%
  summarize(TotalQuantity = sum(Quantity)) %>%
  arrange(desc(TotalQuantity)) %>%
  head(10)

# Plot the top 10 products by quantities
ggplot(product_quantities, aes(x = TotalQuantity, y = reorder(Description, TotalQuantity))) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Create a horizontal bar plot
  labs(title = "Top 10 Products by Quantity",
       x = "Total Quantities",
       y = "Product") +
  theme_minimal()
```

Overlapping Products

```{r}
# Merge product_sales and product_quantities on Description
product_sales_quantity <- merge(
  product_sales,
  product_quantities,
  by = "Description",
  suffixes = c("_sales", "_quantity")
)

# Calculate the percentage of total sales accounted for by the top 10 products
total_sales <- sum(df_no_duplicates$Sales)
top_10_sales <- sum(product_sales$TotalSales)
percentage_top_10_sales <- (top_10_sales / total_sales) * 100
percentage_top_10_sales
```

5.4 Sales by Country

Sales by Country

```{r}
# Group by Country and calculate total sales
sales_by_country <- df_no_duplicates %>%
  group_by(Country) %>%
  summarize(TotalSales = sum(Sales)) %>%
  arrange(desc(TotalSales))

# Summary of sales by country
summary(sales_by_country$TotalSales)

# Filter out negative sales
sales_by_country <- sales_by_country %>%
  filter(TotalSales >= 0)

# Treemap of sales by country
treemap(sales_by_country,
        index = "Country",
        vSize = "TotalSales",
        type = "index",
        title = "Sales by Country (Treemap)",
        fontsize.title = 14,
        fontsize.labels = 12,
        palette = "Blues",
        border.col = "white",
        alpha = 0.8)
```

Top 5 Countries

```{r}
# Get the top 5 countries by sales
top_5_countries <- sales_by_country %>%
  head(5) %>%
  pull(Country)

# Top products by sales for the top 5 countries
top_products_by_country <- df_no_duplicates %>%
  filter(Country %in% top_5_countries) %>%
  group_by(Country, Description) %>%
  summarize(TotalSales = sum(Sales), .groups = 'drop') %>%
  group_by(Country) %>%
  group_modify(~ .x %>% arrange(desc(TotalSales))) %>%
  slice_head(n = 5)

# Plot the top 5 products by sales for the top 5 countries
ggplot(top_products_by_country, aes(x = Description, y = TotalSales, fill = Country)) +
  geom_bar(stat = "identity", position = "dodge") +  # Create grouped bar plot
  labs(title = "Top 5 Products by Sales for Top 5 Countries",
       x = "Product",
       y = "Sales") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

Orders by Country

```{r}
# Count the number of orders by country
orders_by_country <- df_no_duplicates %>%
  count(Country, sort = TRUE)

# Plot the number of orders by country
ggplot(orders_by_country, aes(x = reorder(Country, -n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Orders by Country",
       x = "Country",
       y = "Number of Orders") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The UK accounts for the largest portion of sales, followed by EIRE, the Netherlands, Germany, and France.
Each country has a distinct set of top products by sales, highlighting the potential for personalized marketing.
Orders are positively correlated with sales.


5.5 Customer Behavior
Rename Customer ID Column

```{r}
df_no_duplicates <- df_no_duplicates %>%
  rename(CustomerID = `Customer ID`)
```

Top 10 Customers by Sales

```{r}
# Group by CustomerID and calculate total sales
top_10_customers <- df_no_duplicates %>%
  group_by(CustomerID) %>%
  summarize(TotalSales = sum(Sales)) %>%
  arrange(desc(TotalSales)) %>%
  head(11) %>%
  filter(CustomerID != -1)

# Convert CustomerID to character
top_10_customers <- top_10_customers %>%
  mutate(CustomerID = as.character(CustomerID))

# Filter out CustomerID == "-1" and sort by TotalSales
top_10_customers_filtered <- top_10_customers %>%
  filter(CustomerID != "-1") %>%
  arrange(desc(TotalSales))

# Plot the top 10 customers by total sales
ggplot(top_10_customers_filtered, aes(x = TotalSales, y = reorder(CustomerID, TotalSales))) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Create horizontal bar plot
  labs(title = "Top 10 Customers by Total Sales",
       x = "Total Sales",
       y = "Customer ID") +
  theme_minimal()
```

Top 10 Customers by Sales and Quantities

```{r}
# Group by CustomerID, calculate total sales and quantities
top_10_sales_customers <- df_no_duplicates %>%
  group_by(CustomerID) %>%
  summarize(TotalSales = sum(Sales),
            TotalQuantity = sum(Quantity)) %>%
  arrange(desc(TotalSales)) %>%
  head(10) %>%
  filter(CustomerID != -1)

# Convert CustomerID to character
top_10_sales_customers <- top_10_sales_customers %>%
  mutate(CustomerID = as.character(CustomerID))

# Reshape the data for plotting
top_10_sales_customers_long <- top_10_sales_customers %>%
  pivot_longer(cols = c(TotalSales, TotalQuantity), names_to = "Metric", values_to = "Value")

# Plot the top 10 customers by sales and quantities
ggplot(top_10_sales_customers_long, aes(x = CustomerID, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +  # Create grouped bar plot
  labs(title = "Top 10 Customers by Sales and Quantities",
       x = "Customer ID",
       y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  scale_fill_manual(values = c("TotalSales" = "steelblue", "TotalQuantity" = "orange")) +  # Custom colors
  guides(fill = guide_legend(title = "Metric"))  # Add legend
```

Customer Sales Over Time

```{r}
# Extract YearMonth from InvoiceDate
df_no_duplicates$YearMonth <- format(df_no_duplicates$InvoiceDate, "%Y-%m")

# Filter data for top 10 customers
df_no_duplicates_top_10_customers <- df_no_duplicates[df_no_duplicates$CustomerID %in% top_10_customers$CustomerID, ]

# Group by CustomerID and YearMonth, calculate total sales
customer_sales_over_time <- df_no_duplicates_top_10_customers %>%
  group_by(CustomerID, YearMonth) %>%
  summarise(Sales = sum(Sales), .groups = 'drop')

# Convert YearMonth and CustomerID to character
customer_sales_over_time$YearMonth <- as.character(customer_sales_over_time$YearMonth)
customer_sales_over_time$CustomerID <- as.character(customer_sales_over_time$CustomerID)

# Plot customer sales over time
ggplot(customer_sales_over_time[customer_sales_over_time$CustomerID != "-1", ], aes(x = YearMonth, y = Sales, color = CustomerID, group = CustomerID)) +
  geom_line() +
  labs(title = "Customer Sales Over Time", x = "Year-Month", y = "Total Sales") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Top 10 Customers Summary

```{r}
top_10_customers <- df_no_duplicates_top_10_customers[df_no_duplicates_top_10_customers$CustomerID != -1, ]

head(top_10_customers)
```

```{r}
# Summarize top 10 customers by sales, purchase frequency, and country
top_10_customers_summary <- top_10_customers %>%
  group_by(CustomerID) %>%
  summarise(Sales = sum(Sales), PurchaseFrequency = n(), .groups = 'drop') %>%
  arrange(desc(Sales))

# Get the most frequent country for each customer
top_10_customers_country <- top_10_customers %>%
  group_by(CustomerID) %>%
  summarise(Country = names(sort(table(Country), decreasing = TRUE))[1], .groups = 'drop')

# Merge summaries
top_10_customers_summary <- merge(top_10_customers_summary, top_10_customers_country, by = "CustomerID")
colnames(top_10_customers_summary)[colnames(top_10_customers_summary) == "Invoice"] <- "PurchaseFrequency"
top_10_customers_summary
```

We identified the top 10 customers, and found a positive correlation between their purchasing power and quantities.
Throughout the year, the top customers do not exhibit any clear seasonal purchasing patterns.
Eight of the top 10 customers are from the UK, with the second-largest customer hailing from the Netherlands.

6. Customer Segmentation - Clustering

Calculate Recency

```{r}
recent_date <- max(df_no_duplicates$InvoiceDate)
df_no_duplicates$Recency <- as.numeric(difftime(recent_date, df_no_duplicates$InvoiceDate, units = "days"))
```

RFM Analysis

```{r}
# Calculate Recency, Frequency, and MonetaryValue for each customer
customer_rfm <- df_no_duplicates %>%
  group_by(CustomerID) %>%
  summarise(
    Recency = as.numeric(difftime(recent_date, max(InvoiceDate), units = "days")),
    Frequency = n_distinct(Invoice),
    MonetaryValue = sum(Sales),
    .groups = 'drop'
  ) %>%
  filter(CustomerID != -1)

# Calculate quantiles for Recency, Frequency, and MonetaryValue
quantiles <- quantile(customer_rfm$Recency, probs = c(0.25, 0.5, 0.75))
quantiles_dict <- list(
  Recency = quantile(customer_rfm$Recency, probs = c(0.25, 0.5, 0.75)),
  Frequency = quantile(customer_rfm$Frequency, probs = c(0.25, 0.5, 0.75)),
  MonetaryValue = quantile(customer_rfm$MonetaryValue, probs = c(0.25, 0.5, 0.75))
)

# Define scoring functions
RScore <- function(x, p, d) {
  if (x <= d[[p]][1]) {
    return(4)
  } else if (x <= d[[p]][2]) {
    return(3)
  } else if (x <= d[[p]][3]) {
    return(2)
  } else {
    return(1)
  }
}

FMScore <- function(x, p, d) {
  if (x <= d[[p]][1]) {
    return(1)
  } else if (x <= d[[p]][2]) {
    return(2)
  } else if (x <= d[[p]][3]) {
    return(3)
  } else {
    return(4)
  }
}

# Assign RFM scores
customer_rfm$R <- sapply(customer_rfm$Recency, RScore, p = "Recency", d = quantiles_dict)
customer_rfm$F <- sapply(customer_rfm$Frequency, FMScore, p = "Frequency", d = quantiles_dict)
customer_rfm$M <- sapply(customer_rfm$MonetaryValue, FMScore, p = "MonetaryValue", d = quantiles_dict)

# Combine RFM scores into a single RFM group and score
customer_rfm$RFMGroup <- paste0(customer_rfm$R, customer_rfm$F, customer_rfm$M)
customer_rfm$RFMScore <- rowSums(customer_rfm[, c("R", "F", "M")])

# Prepare data for clustering
rfm_data <- customer_rfm[, c("Recency", "Frequency", "MonetaryValue")]
scaled_rfm_data <- scale(rfm_data)  # Scale the data
```

Elbow Method for Optimal Clusters

```{r}
# Calculate inertia for different values of k
inertia <- numeric(10)
for (i in 1:10) {
  kmeans_result <- kmeans(scaled_rfm_data, centers = i, nstart = 25)
  inertia[i] <- kmeans_result$tot.withinss
}

# Plot the elbow method
ggplot(data.frame(k = 1:10, inertia = inertia), aes(x = k, y = inertia)) +
  geom_line() +
  geom_point() +
  labs(title = "Elbow Method for Optimal k", x = "Number of Clusters (k)", y = "Inertia") +
  theme_minimal()
```

K-Means Clustering

```{r}
set.seed(42)  # For reproducibility
kmeans_result <- kmeans(scaled_rfm_data, centers = 3, nstart = 25)
customer_rfm$Cluster <- kmeans_result$cluster
```

Cluster Analysis

```{r}
# Summarize cluster characteristics
cluster_analysis <- customer_rfm %>%
  group_by(Cluster) %>%
  summarise(
    Recency_mean = mean(Recency),
    Recency_median = median(Recency),
    Recency_min = min(Recency),
    Recency_max = max(Recency),
    Frequency_mean = mean(Frequency),
    Frequency_median = median(Frequency),
    Frequency_min = min(Frequency),
    Frequency_max = max(Frequency),
    MonetaryValue_mean = mean(MonetaryValue),
    MonetaryValue_median = median(MonetaryValue),
    MonetaryValue_min = min(MonetaryValue),
    MonetaryValue_max = max(MonetaryValue),
    .groups = 'drop'
  )

print(cluster_analysis)
```

Visualize Clusters

```{r}
# Recency Distribution by Cluster
plot1 <- ggplot(customer_rfm, aes(x = factor(Cluster), y = Recency)) +
  geom_boxplot() +
  labs(title = "Recency Distribution by Cluster", x = "Cluster", y = "Recency")

# Frequency Distribution by Cluster
plot2 <- ggplot(customer_rfm, aes(x = factor(Cluster), y = Frequency)) +
  geom_boxplot() +
  labs(title = "Frequency Distribution by Cluster", x = "Cluster", y = "Frequency")

# MonetaryValue Distribution by Cluster
plot3 <- ggplot(customer_rfm, aes(x = factor(Cluster), y = MonetaryValue)) +
  geom_boxplot() +
  labs(title = "MonetaryValue Distribution by Cluster", x = "Cluster", y = "MonetaryValue")

# Combine plots
library(gridExtra)
grid.arrange(plot1, plot2, plot3, ncol = 3)
```

Scatterplots for Clusters

```{r}
# Recency vs MonetaryValue
ggplot(customer_rfm, aes(x = Recency, y = MonetaryValue, color = factor(Cluster))) +
  geom_point() +
  scale_color_viridis_d() +
  labs(title = "Customer Segmentation by Recency and MonetaryValue", x = "Recency", y = "MonetaryValue") +
  theme_minimal()

# Frequency vs MonetaryValue
ggplot(customer_rfm, aes(x = Frequency, y = MonetaryValue, color = factor(Cluster))) +
  geom_point() +
  scale_color_viridis_d() +
  labs(title = "Customer Segmentation by Frequency and MonetaryValue", x = "Frequency", y = "MonetaryValue") +
  theme_minimal()

# Recency vs Frequency
ggplot(customer_rfm, aes(x = Recency, y = Frequency, color = factor(Cluster))) +
  geom_point() +
  scale_color_viridis_d() +
  labs(title = "Customer Segmentation by Recency and Frequency", x = "Recency", y = "Frequency") +
  theme_minimal()
```

Cluster Distribution

```{r}
# Count the number of customers in each cluster
cluster_counts <- table(customer_rfm$Cluster)

# Create a data frame for the pie chart
pie_data <- data.frame(Cluster = names(cluster_counts), Count = as.numeric(cluster_counts))

# Calculate percentages
pie_data$Percentage <- pie_data$Count / sum(pie_data$Count) * 100

# Pie chart
ggplot(pie_data, aes(x = "", y = Count, fill = Cluster)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Distribution of Customer Clusters", fill = "Cluster") +
  theme_void() +
  theme(legend.position = "right")
```

Sales Trends by Cluster

```{r}
# Merge and group by Cluster and YearMonth
cluster_sales_over_time <- df_no_duplicates %>%
  left_join(customer_rfm, by = "CustomerID") %>%
  group_by(Cluster, YearMonth) %>%
  summarise(Sales = sum(Sales), .groups = 'drop')

# Convert YearMonth to timestamp
cluster_sales_over_time$YearMonth <- as.Date(paste0(cluster_sales_over_time$YearMonth, "-01"))

# Filter the data to only include 3 clusters
cluster_sales_over_time_filtered <- cluster_sales_over_time %>%
  filter(Cluster %in% c(1, 2, 3))

# Plot the sales trends for only 3 clusters
ggplot(cluster_sales_over_time_filtered, aes(x = YearMonth, y = Sales, color = factor(Cluster))) +
  geom_line() +
  labs(title = "Cluster Sales Over Time", x = "Year-Month", y = "Total Sales") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Top Products by Cluster

```{r}
# Group by Cluster and Description, calculate total sales
top_products_by_cluster <- df_no_duplicates %>%
  left_join(customer_rfm, by = "CustomerID") %>%
  group_by(Cluster, Description) %>%
  summarise(Sales = sum(Sales), .groups = "drop") %>%
  group_by(Cluster) %>%
  arrange(Cluster, desc(Sales)) %>%
  slice_head(n = 5) %>%
  ungroup()

# Filter the data to only include 3 clusters
top_products_by_cluster <- top_products_by_cluster %>%
  filter(Cluster %in% c(1, 2, 3))

# Plotting the top products per cluster
ggplot(top_products_by_cluster, aes(x = Description, y = Sales, fill = factor(Cluster))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 5 Products by Sales for each Cluster", x = "Product", y = "Sales") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

There are three clusters. Cluster 2 has the highest monetary value with the shortest recency, but the least frequency. Cluster 1 has the lowest monetary value, the longest recency, and an average frequency of 1, indicating only one purchase. Interestingly, this cluster ceased purchasing in July 2020. Further investigation is needed to understand the reasons behind this.

The high-value Cluster 2 maintains a stable purchasing pattern through the year, while Cluster 1 shows a seasonal increase starting in August 2010, peaking in November 2010. Although their average purchase amount is not the highest, their total spend accounts for the majority of sales.

We also found that the three clusters have different product preferences, which highlights the potential for targeted marketing.


